\documentclass{article}


% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2025


% ready for submission
%\usepackage{neurips_2025}


% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2025}


% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final, nonatbib]{neurips_2025}


% to avoid loading the natbib package, add option nonatbib:
%\usepackage[nonatbib]{neurips_2025}
\usepackage{makecell}

\usepackage{pifont} % for ding symbols
\newcommand{\xmark}{\ding{55}} % define xmark as a cross

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
%\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
%\usepackage{xcolor}         % colors
\usepackage{xcolor} % Make sure this is in your preamble


%%%%%%%%%%%%%%% added by me

\usepackage{xspace}
\usepackage{soul}  
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\mycode}[1]{{\color{orange}#1}}
\newcommand{\todo}[1]{{\color{red}#1}}
\newcommand{\TODO}[1]{\textbf{\color{red}[TODO: #1]}}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{adjustbox}
\def\onedot{.\xspace}
\def\eg{\emph{e.g}\onedot} \def\Eg{\emph{E.g}\onedot}
\def\ie{\emph{i.e}\onedot} \def\Ie{\emph{I.e}\onedot}
\usepackage{mathtools}
\usepackage{textcomp}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{color, colortbl}
\usepackage{makecell}
\usepackage[dvipsnames]{xcolor}
\usepackage{tabulary}
\usepackage{subcaption}
\usepackage{wrapfig}

\definecolor{citecolor}{RGB}{0, 113, 188}
\usepackage[pagebackref,breaklinks,colorlinks,citecolor=citecolor,linkcolor=red]{hyperref}
\usepackage[capitalize]{cleveref}
\crefname{section}{Sec.}{Secs.}
\Crefname{section}{Section}{Sections}
\Crefname{table}{Table}{Tables}
\crefname{table}{Tab.}{Tabs.}
\newcommand{\pub}[1]{\color{gray}{\scriptsize{[{#1}]}}}
\newcommand{\puB}[1]{\color{gray}{\small{{#1}}}}
\newcommand{\myparagraph}[1]{\vspace{4pt}\noindent\textbf{#1.}}
\newcommand{\myparagraphnodot}[1]{\vspace{4pt}\noindent\textbf{#1}}
\definecolor{mylightgray}{gray}{0.4}
\definecolor{mylightgrayours}{gray}{0.2}
\definecolor{lightblue}{rgb}{0.8, 0.8, 1.0}


\newcommand\ours{SANSA}
\newcommand\sam{\textsc{SAM2}}

%%%%%%%%%%%%%%%

\title{SANSA: Unleashing the Hidden Semantics in \\ SAM2 for Few-Shot Segmentation}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{Claudia Cuttano${^{*} }$ \quad Gabriele Trivigno\thanks{Equal contribution.}
\quad
Giuseppe Averta
\quad
Carlo Masone  \\[0.2cm]
Politecnico di Torino \\
{\tt\small \{name.surname\}@polito.it \
}
}

\begin{document}


\maketitle


\begin{abstract}
Few-shot segmentation aims to segment unseen categories from just a handful of annotated examples. This requires mechanisms to identify semantically related objects across images and accurately produce masks. We note that Segment Anything 2 (SAM2), with its \textit{prompt-and-propagate} mechanism, provides strong segmentation capabilities and a built-in feature matching process. However, we show that its representations are entangled with task-specific cues optimized for object tracking, which impairs its use for tasks requiring higher level semantic understanding. Our key insight is that, despite its class-agnostic pretraining, SAM2 already encodes rich semantic structure in its features. We propose \textbf{\ours{}} (\textbf{S}emantically \textbf{A}lig\textbf{N}ed \textbf{S}egment\textbf{A}nything 2), a framework that makes this latent structure explicit, and repurposes SAM2 for few-shot segmentation through minimal task-specific modifications. SANSA achieves state-of-the-art on few-shot segmentation benchmarks designed to assess generalization and outperforms generalist methods in the popular in-context setting. Additionally, it supports  flexible promptable interaction via points, boxes, or scribbles, and remains significantly faster and more compact than prior approaches.  Code at: \url{https://github.com/ClaudiaCuttano/SANSA}.
\end{abstract}


\input{sec/1_intro}
\input{sec/2_related}
\input{sec/3_method}
\input{sec/4_esperiments}
\input{sec/5_conclusions}

\myparagraph{Acknowledgements}
Claudia Cuttano was supported by the Sustainable Mobility Center (CNMS) which received funding from the European Union Next Generation EU (Piano Nazionale di Ripresa e Resilienza (PNRR), Missione 4 Componente 2 Investimento 1.4 "Potenziamento strutture di ricerca e creazione di "campioni nazionali di R\&S" su alcune Key Enabling Technologies") with grant agreement no. CN\_00000023. 

Gabriele Trivigno, Giuseppe Averta and Carlo Masone were supported by FAIR - Future Artificial Intelligence Research which received funding from the European Union Next-GenerationEU (PIANO NAZIONALE DI RIPRESA E RESILIENZA (PNRR) – MISSIONE 4 COMPONENTE 2, INVESTIMENTO 1.3 – D.D. 1555 11/10/2022, PE00000013).  This manuscript reflects only the authors’ views and opinions, neither the European Union nor the European Commission can be considered responsible for them.

We acknowledge the CINECA award
under the ISCRA initiative, for the availability of high performance computing resources.

%\input{sec/X_checklist}
{
    \small
    \bibliographystyle{abbrv}
    \bibliography{main}
}

\newpage

\input{sec/6_supp}

\newpage

\end{document}